---
output: rmarkdown::github_document
---

## debiasedmcmc

This package contains scripts to reproduce the results of the arXiv report "Unbiased Markov chain Monte Carlo with couplings", by Pierre E. Jacob, John O'Leary and Yves F Atchade.

This is not a general-purpose statistical software, but a collection of scripts intended to reproduce figures and tables. Use at your own risk!!

The Polya-Gamma sampler is taken from the package BayesLogit of Nick Polson, James Scott, and Jesse Windle:

https://cran.r-project.org/src/contrib/Archive/BayesLogit/

That package can be downloaded and installed with "R CMD INSTALL". Or in R via

```{r installbayeslogit, eval=FALSE, echo=TRUE}
packageurl <- "https://cran.r-project.org/src/contrib/Archive/BayesLogit/BayesLogit_0.6.tar.gz"

install.packages(packageurl, repos=NULL, type="source")
```

Note that the above requires gfortran to compile. For instance on Mac OS X gfortran can be found on https://github.com/fxcoudert/gfortran-for-macOS/releases.

The folder inst/reproduce/ contains the scripts to reproduce the figures. Each sub-folder has a "run all" script to run all the scripts in the correct order.
The folder inst/check/ contains internal checks, and vignettes/ contains tutorials on
how to use the package's main functions.

### Installation

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("pierrejacob/debiasedmcmc")
```

Additionally you might want to install other packages, to help with parallel computation:
```{r installparallel, eval=FALSE}
install.packages(c("doParallel", "doRNG"))
```
and to help with manipulating results and plotting:
```{r installplot, eval=FALSE}
install.packages(c("dplyr", "tidyr", "ggplot2", "viridis"))
```

### Usage
